---
title: "Statistical Learning Final Project"
subtitle: "Employee Attrition Classification"

author: [Zeynep TUTAR - 2106038, Aysenur Oya ÖZEN - 0000000]

date: "09-07-2024"
toc: true
toc-own-page: true
number_sections: true
df_print: kable
fontsize: 11pt
lang: en-GB
titlepage: true
titlepage-background: "img/unipdbg.pdf"
header-center: "\\leftmark"
header-right: "\\thetitle"
footer-left: "\\hspace{1cm}"
footer-center: "Page \\thepage"
footer-right: "\\hspace{1cm}"
hyperrefoptions:
- linktoc=all
biblio-style: apsr
colorlinks: true

header-includes:
   - \usepackage{setspace}
   - \newpage
   - \usepackage{hyperref}
   - \usepackage{multirow}
   - \usepackage{amsmath}
   - \usepackage{algorithm}
   - \usepackage{algorithmic}
   - \usepackage{authblk}
   - \usepackage{graphicx}

output:
  pdf_document:
    template: "template/eisvogel.tex"
    citation_package: natbib
    keep_tex: false
    fig_caption: true
    latex_engine: pdflatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      warning = FALSE, message = FALSE,
                      tidy.opts = list(width.cutoff = 55), 
                      tidy = TRUE, comment = '',
                      fig.path='figs/',
                      cache = TRUE, cache.path = '_cache/')
```

# Introduction to Dataset

The aim of this project is to develop two predictive models to determine employee attrition of a company. The dataset[^1] used for this project is a simulated dataset designed for the analysis and prediction of employee attrition. It contains detailed information about various aspects of an employee's profile, including demographics, job-related features, and personal circumstances.The dataset contains 74,498 samples. Each record includes a unique Employee ID and features that influence employee attrition. The goal is to understand the factors contributing to attrition and develop predictive models to identify at-risk employees.

[^1]: <https://www.kaggle.com/datasets/stealthtechnologies/employee-attrition-dataset/data>

The dataset is already splitted into train and test but in order to better understand the data, it is crucial to analyse the dataset as a whole.

```{r import_data}

# import the train and test datasets
data_train <- read.csv("data/train.csv")
data_test <- read.csv("data/test.csv")

# merge the datasets
data <- rbind(data_train, data_test)
attach(data)
```

## Description of the Features

The features of the dataset are presented below:

-   **Employee ID:** A unique identifier assigned to each employee.

-   **Age:** The age of the employee, ranging from 18 to 60 years.

-   **Gender:** The gender of the employee

-   **Years at Company:** The number of years the employee has been working at the company.

-   **Monthly Income:** The monthly salary of the employee, in dollars.

-   **Job Role:** The department or role the employee works in, encoded into categories such as Finance, Healthcare, Technology, Education, and Media.

-   **Work-Life Balance:** The employee's perceived balance between work and personal life, (Poor, Below Average, Good, Excellent)

-   **Job Satisfaction:** The employee's satisfaction with their job: (Very Low, Low, Medium, High)

-   **Performance Rating:** The employee's performance rating: (Low, Below Average, Average, High)

-   **Number of Promotions:** The total number of promotions the employee has received.

-   **Distance from Home:** The distance between the employee's home and workplace, in miles.

-   **Education Level:** The highest education level attained by the employee: (High School, Associate Degree, Bachelor’s Degree, Master’s Degree, PhD)

-   **Marital Status:** The marital status of the employee: (Divorced, Married, Single)

-   **Job Level:** The job level of the employee: (Entry, Mid, Senior)

-   **Company Size:** The size of the company the employee works for: (Small,Medium,Large)

-   **Company Tenure:** The total number of years the employee has been working in the industry.

-   **Remote Work:** Whether the employee works remotely: (Yes or No)

-   **Leadership Opportunities:** Whether the employee has leadership opportunities: (Yes or No)

-   **Innovation Opportunities:** Whether the employee has opportunities for innovation: (Yes or No)

-   **Company Reputation:** The employee's perception of the company's reputation: (Very Poor, Poor,Good, Excellent)

-   **Employee Recognition:** The level of recognition the employee receives:(Very Low, Low, Medium, High)

-   **Attrition:** Whether the employee has left the company, encoded as 0 (stayed) and 1 (Left).

# Data Analysis

In order to develop predictive models, first it is necessary to perform exploratory data analysis (EDA) and modify the format of the data if necessary.

```{r EDA}
# first column contains Employee IDs, so not necessary for summary
summary(data[,-1], )
```

## Data Preprocessing

To prepare the dataset for further analysis, several data preprocessing steps are performed:

1.  Converting categorical features to factors
2.  Removing features
3.  Handling na values
4.  etc...

```{r}
#EDA

```

### Outliers

```{r}
#EDA

```

### Visualization

```{r}
#EDA

```

As a result of the analysis, the following observations were made regarding the characteristics of the data:

## Features vs. Target

### Categorical Features vs. Target

### Numerical Features vs. Target

## Correlation Matrix

### Partial Correlation Matrices

# Data Preparation

## Categorical to Numerical Feature Conversion

## Train-Test-Split

## Feature and Output Samples

# Predictive Classification Models

## Logistic Regression

### Basic Logistic Classifier

### Logistic Regression with Backward Variable Selection

### Logistic Regression with Shrinkage Method

### ROC Curve & Comparison of Logistic Classifiers

## Another Classification Model

# Model Results

## Performance Metrics and Confusion Matrix

# Conclusion
